---
title: "Understanding and Resolving Duplicate Records in ESSENCE Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding and Resolving Duplicate Records in ESSENCE Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(sysPrep)
library(dplyr)
```

## Why ESSENCE Contains Duplicate Records

The vast majority of ESSENCE records are updated correctly via HL7 update
messages, which are accurately appended to the existing record without
generating additional rows. Duplicate records are the exception, not the rule.
When duplicates do occur, they arise through distinct mechanisms with different
causes and different implications for how deduplication should be approached.
Without deduplication, these additional rows inflate case counts, distort trend
analyses, and produce inaccurate rates.

**Why duplication happens.** BioSense, the NSSP platform underlying ESSENCE,
computes a record identifier -- `C_BioSense_ID` (called EssenceID in the
ESSENCE backend) -- from a small set of calculated fields: most often
`C_Visit_Date_Time` and `C_Unique_Patient_ID`, and `C_BioSense_Facility_ID`
(`Hospital` in this package and in `dataDetails` API pulls) when a transfer
is involved. As long as those calculated fields are transmitted consistently
for a given encounter, updates to other fields -- a discharge disposition, a
lab result, a diagnosis code -- are correctly appended to the existing
record, with no additional row created. This is why the vast majority of
ESSENCE updates do not produce duplicates. NSSP is not the source of the
problem: a new `C_BioSense_ID` is computed, and an apparent second record
created, only when one of the calculated fields disagrees between
transmissions of what should be the same encounter -- most often
`C_Visit_Date_Time` or `C_Unique_Patient_ID`, and less often
`C_BioSense_Facility_ID` during a transfer. BioSense has no way to tell
whether such a change is a genuine correction or an unwarranted, feed-level
error; either way, the visit is duplicated. When a site transmits fully
accurate calculated fields for every visit, duplication from this mechanism
is essentially zero. The two most common ways it occurs in ESSENCE data are:

**`visit_date_change`.** `C_BioSense_ID` is derived from `C_Visit_Date` and
`C_Visit_Date_Time`, which are frequently populated from `Admit_Date_Time`.
Some hospitals' systems update `C_Visit_Date_Time` as providers interact
with the patient over the course of the visit, rather than fixing it at
initial registration. When one of those updates crosses midnight --
changing what was a pre-midnight value to a post-midnight value -- NSSP
computes a new `C_BioSense_ID` for the same `Visit_ID`, producing two rows
that appear to represent different BioSense records but refer to the same
encounter. This is the most widespread duplication type across NSSP sites,
though the exact scope varies by site and facility, and the affected
proportion of visits is typically small. Many hospitals handle
midnight-crossing visits correctly by not modifying `Admit_Date_Time` or
`C_Visit_Date_Time` after the patient's initial registration; the duplicate
arises only at facilities whose systems continue to update it afterward.

The impact of this duplication is disproportionate in small-count contexts. A
syndrome definition for a relatively rare condition -- such as a specific
overdose agent or an emerging chief complaint pattern -- may produce only a
handful of true encounters at a given facility on a given day. If those
encounters generate multiple `C_BioSense_ID` values, the apparent count
presented to an analyst or flagged by an anomaly detection algorithm can be
substantially higher than the true count, potentially triggering alerts that
do not reflect genuine changes in incidence. This is the clinical and
surveillance consequence that motivates deduplication even when the affected
proportion of total visits is low.

**`pid_change`.** `C_BioSense_ID` also incorporates `C_Unique_Patient_ID`
(which maps to medical record number in Kentucky). A patient may be
registered under a temporary identifier at arrival and assigned their
permanent MRN after identity verification; the facility submits an update
message with the corrected `C_Unique_Patient_ID` while `Visit_ID` itself
stays the same. Because BioSense recomputes `C_BioSense_ID` from the new
`C_Unique_Patient_ID`, the update produces an additional row rather than
correcting the existing one -- the original row, tied to the earlier
`C_Unique_Patient_ID`, is left in place and still reflects the earlier
state of the visit, while the new row carries the corrected identifier and
current information. Each identifier change produces one additional row
for the same `Visit_ID`.

**`patient_class_change`.** Under normal HL7 processing, patient class
transitions during a visit -- for example, from emergency department to
inpatient admission -- are handled without generating duplicate rows: the
transition is appended to `C_Patient_Class_List`, `C_Patient_Class` is
updated, and the `HasBeenE`, `HasBeenI`, `HasBeenAdmitted` flags (ED-visit
and inpatient-admission indicator columns) are revised in place. A `patient_class_change` duplicate appears only when something
additional causes a new row to be submitted for the same `Visit_ID` coincident
with a class transition -- such as a feed configuration issue or a concurrent
`visit_date_change`. This mechanism requires `c_patient_class` to be present
in the data pull. See also `vignette("encounter-linkage")`, which addresses
multi-class visits through encounter linkage rather than deduplication.

**Cross-pull double-counting when combining ED and direct-admit data.**
The mechanisms above all produce duplicate rows *within* a single pull,
which `summarize_duplicates()`, `classify_duplicates()`, and `dedupe()`
are built to detect and resolve. A related but distinct risk arises when
combining a `HasBeenE = 1` pull with a separately queried
`HasBeenAdmitted = 1` pull to build a combined "severe visit" count:
pre-aggregated ESSENCE data (timeSeries, tableBuilder) cannot distinguish
a true direct admission from a mis-submitted one whose corresponding ED
record incorrectly shows `HasBeenAdmitted = 0`. Combining the two pulls
without linking them double-counts the mis-submitted cases. `dedupe()` is
not the right tool for this: even though the ED and mis-submitted
admission records share a `facility_col x visit_col` key, they carry
complementary, not redundant, information -- the admission record's
discharge disposition and outcome are not present on the ED record -- so
discarding one (as `dedupe()` does) silently loses information that
merging (as `link_encounters()` does) preserves. See
`vignette("encounter-linkage")` and `link_encounters()` for the dedicated
fix.

## Quantifying Duplicates: `summarize_duplicates()`

Before deduplicating, call `summarize_duplicates()` to understand the scope
and distribution of duplicates in your data. This step is diagnostic and does
not alter the data.

```{r summarize}
dups <- summarize_duplicates(essence_raw)
dups
```

The printed output shows dataset-wide duplicate counts and a per-facility
breakdown. Accessing the list components directly provides the underlying
data:

```{r summarize-components}
# Dataset-level counts
dups$overall

# Per-facility counts (facilities with duplicates only)
dups$by_facility
```

`$overall` gives the total number of duplicated `facility × Visit_ID` pairs and
the proportion of unique visits affected. `$by_facility` gives per-facility
counts of duplicated visit identifiers, sorted by total duplicated visits
descending. A facility with a disproportionately high duplicate rate may
indicate a feed configuration issue -- for example, a registration system
that routinely updates `C_Visit_Date_Time` as providers interact with the
patient (producing `visit_date_change` duplicates whenever an update
crosses midnight), or a workflow that changes `C_Unique_Patient_ID` between
update messages for the same visit (producing `pid_change` duplicates).

> **Why is the deduplication key `HospitalName × Visit_ID`, not `Visit_ID` alone?**
> The same `Visit_ID` may appear at two facilities if a patient is transferred
> mid-visit and both facilities transmit records with their own internal visit
> identifiers that happen to be numerically identical. Deduplicating by
> `Visit_ID` alone would incorrectly collapse these into a single encounter.
> `sysPrep` always deduplicates within facility.

## Understanding the Mechanism: `classify_duplicates()`

After quantifying duplicates, classify their mechanism to inform your
deduplication strategy and identify potential feed-level issues.

```{r classify}
classified <- classify_duplicates(essence_raw)
classified
```

```{r classify-overall}
# Type distribution across the full dataset
classified$overall
```

```{r classify-facility}
# Per-facility breakdown by type (wide format)
classified$by_facility
```

The `$overall` component shows how many duplicate groups belong to each
mechanism type. The `$by_facility` component shows the distribution of
mechanisms per facility in wide format -- useful for identifying facilities
whose duplicates are concentrated in a particular mechanism.

When `c_patient_class` is not included in your ESSENCE pull, the function
issues an informational message and skips `patient_class_change` detection.
All other mechanism types remain fully functional.

> **Why classify before removing duplicates?**
> Classification reveals whether your duplicates are dominated by retransmissions
> (suggesting `keep = "last"` is appropriate) or by `visit_date_change` events
> (which may warrant examining whether the date change affects your analysis
> window). It also identifies whether a specific facility is generating an
> unusual volume of a particular duplication type -- which can point to a
> feed-level configuration problem worth reporting to the facility or NSSP.

For row-level analysis, use `return_format = "tibble"` to join classifications
back to the original data:

```{r classify-tibble}
# Join mechanism type back to raw data for row-level inspection.
# classify_duplicates() returns cleaned (snake_case) column names,
# so clean essence_raw first to align join keys.
typed <- essence_raw |>
  janitor::clean_names() |>
  dplyr::left_join(
    classify_duplicates(essence_raw, return_format = "tibble"),
    by = c("hospital_name", "visit_id")
  )

dplyr::count(typed, dup_type)
```

## Removing Duplicates: `dedupe()`

With duplicates characterized, remove them using `dedupe()`. The function
retains exactly one row per `facility × visit_col` group.

```{r dedupe}
deduped <- dedupe(essence_raw, order_by = Arrived_Date_Time, keep = "last")

cat("Rows before: ", nrow(essence_raw), "\n")
cat("Rows after:  ", nrow(deduped), "\n")
cat("Rows removed:", nrow(essence_raw) - nrow(deduped), "\n")
```

```{r verify-zero}
# Confirm zero duplicate groups remain
summarize_duplicates(deduped)$overall
```

The `order_by` argument specifies which column to use for ranking rows within
each duplicate group. `Arrived_Date_Time` -- the timestamp when NSSP received
the record -- is the recommended ordering column for production surveillance
because the most recently received row is most likely to reflect updated
clinical information.

## Choosing a Deduplication Strategy

The appropriate strategy depends on your surveillance context:

| Scenario | Recommended strategy |
|----------|---------------------|
| Rolling surveillance (production monitoring) | `order_by = Arrived_Date_Time, keep = "last"` |
| Fixed-window retrospective analysis | `order_by = Arrived_Date_Time, keep = "first"` |
| Field completeness matters (e.g., maximizing diagnosis code coverage) | `keep = "most_complete"` |
| Unsure -- inspect before deciding | Call `classify_duplicates()` first |

**`keep = "last"`** retains the row with the highest value of `order_by` -- the
most recently transmitted record. Appropriate when the latest transmission is
most current and complete, which is true for most retransmission scenarios.

**`keep = "first"`** retains the row with the lowest value of `order_by` --
the original transmission. Appropriate for fixed-window analyses where you want
to capture the visit as it was first reported, without post-hoc field updates.

**`keep = "most_complete"`** retains the row with the fewest `NA` values across
all columns. Appropriate when no single transmission is reliably more current
than another and field completeness is the primary concern -- for example, when
discharge diagnosis codes are populated inconsistently across retransmissions.

When the `order_by` column itself contains ties (multiple rows with the same
`Arrived_Date_Time`), `dedupe()` retains the first occurrence among tied rows
and issues an informational message. To avoid tie-breaking ambiguity, use
`C_BioSense_ID` as a secondary sort or combine `keep = "most_complete"` with
an explicit `order_by`.

## Next Steps

- `vignette("encounter-linkage")`: After deduplication, understand how
  `link_encounters()` handles the structural exclusion of direct admissions
  from `HasBeenE = 1` filtered queries.

- `vignette("getting-started")`: The full recommended pipeline, combining
  deduplication with care setting filtering and geographic attribution.

- `?dedupe`, `?summarize_duplicates`, `?classify_duplicates`: Full parameter
  documentation and additional examples.
