---
title: "Understanding and Resolving Duplicate Records in ESSENCE Data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Understanding and Resolving Duplicate Records in ESSENCE Data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(sysPrep)
library(dplyr)
```

## Why ESSENCE Contains Duplicate Records

The vast majority of ESSENCE records are updated correctly via HL7 update
messages, which are accurately appended to the existing record without
generating additional rows. Duplicate records are the exception, not the rule.
When duplicates do occur, they arise through distinct mechanisms with different
causes and different implications for how deduplication should be approached.
Without deduplication, these additional rows inflate case counts, distort trend
analyses, and produce inaccurate rates.

**Standard retransmission.** A facility resubmits a record after updating
clinical fields -- adding a discharge disposition, a lab result, or a diagnosis
code. The National Syndromic Surveillance Program (NSSP) appends the new
transmission rather than overwriting the original, resulting in two rows with
the same `HospitalName × Visit_ID` combination. The newer row typically
contains more complete information. Retaining the most recently transmitted row
(`order_by = Arrived_Date_Time, keep = "last"`) is appropriate for most
surveillance contexts.

**`visit_date_change`.** `C_BioSense_ID` is derived from `C_Visit_Date` and
`C_Visit_Date_Time`, which are frequently populated from `Admit_Date_Time`.
When a hospital treats `C_Visit_Date_Time` as a modifiable field and submits
an update that crosses midnight -- changing what was a pre-midnight value to a
post-midnight value -- NSSP computes a new `C_BioSense_ID` for the same
`Visit_ID`, producing two rows that appear to represent different BioSense
records but refer to the same encounter. This is the most widespread
duplication type across NSSP sites, though the exact scope varies by site and
facility, and the affected proportion of visits is typically small. Many
hospitals handle midnight-crossing visits correctly by not modifying
`Admit_Date_Time` or `C_Visit_Date_Time` after the patient's initial
registration; the duplicate arises only at facilities that treat these fields
as modifiable.

The impact of this duplication is disproportionate in small-count contexts. A
syndrome definition for a relatively rare condition -- such as a specific
overdose agent or an emerging chief complaint pattern -- may produce only a
handful of true encounters at a given facility on a given day. If those
encounters generate multiple `C_BioSense_ID` values, the apparent count
presented to an analyst or flagged by an anomaly detection algorithm can be
substantially higher than the true count, potentially triggering alerts that
do not reflect genuine changes in incidence. This is the clinical and
surveillance consequence that motivates deduplication even when the affected
proportion of total visits is low.

**`pid_change`.** The patient identifier (`C_Unique_Patient_ID`, which maps to
medical record number in Kentucky) was corrected or updated mid-visit by the
facility. A patient may be registered under a temporary identifier at arrival
and assigned their permanent MRN after identity verification. Each identifier
change produces an additional row for the same `Visit_ID`.

**`patient_class_change`.** Under normal HL7 processing, patient class
transitions during a visit -- for example, from emergency department to
inpatient admission -- are handled without generating duplicate rows: the
transition is appended to `C_Patient_Class_List`, `C_Patient_Class` is
updated, and the `HasBeenE`, `HasBeenI`, `HasBeenAdmitted` flags (ED-visit
and inpatient-admission indicator columns) are revised in place. A `patient_class_change` duplicate appears only when something
additional causes a new row to be submitted for the same `Visit_ID` coincident
with a class transition -- such as a feed configuration issue or a concurrent
`visit_date_change`. This mechanism requires `c_patient_class` to be present
in the data pull. See also `vignette("encounter-linkage")`, which addresses
multi-class visits through encounter linkage rather than deduplication.

## Quantifying Duplicates: `summarize_duplicates()`

Before deduplicating, call `summarize_duplicates()` to understand the scope
and distribution of duplicates in your data. This step is diagnostic and does
not alter the data.

```{r summarize}
dups <- summarize_duplicates(essence_raw)
dups
```

The printed output shows dataset-wide duplicate counts and a per-facility
breakdown. Accessing the list components directly provides the underlying
data:

```{r summarize-components}
# Dataset-level counts
dups$overall

# Per-facility counts (facilities with duplicates only)
dups$by_facility
```

`$overall` gives the total number of duplicated `facility × Visit_ID` pairs and
the proportion of unique visits affected. `$by_facility` gives per-facility
counts of duplicated visit identifiers, sorted by total duplicated visits
descending. A facility with a disproportionately high duplicate rate may
indicate a feed configuration issue -- for example, a facility that does not
suppress retransmission for unchanged records.

> **Why is the deduplication key `HospitalName × Visit_ID`, not `Visit_ID` alone?**
> The same `Visit_ID` may appear at two facilities if a patient is transferred
> mid-visit and both facilities transmit records with their own internal visit
> identifiers that happen to be numerically identical. Deduplicating by
> `Visit_ID` alone would incorrectly collapse these into a single encounter.
> `sysPrep` always deduplicates within facility.

## Understanding the Mechanism: `classify_duplicates()`

After quantifying duplicates, classify their mechanism to inform your
deduplication strategy and identify potential feed-level issues.

```{r classify}
classified <- classify_duplicates(essence_raw)
classified
```

```{r classify-overall}
# Type distribution across the full dataset
classified$overall
```

```{r classify-facility}
# Per-facility breakdown by type (wide format)
classified$by_facility
```

The `$overall` component shows how many duplicate groups belong to each
mechanism type. The `$by_facility` component shows the distribution of
mechanisms per facility in wide format -- useful for identifying facilities
whose duplicates are concentrated in a particular mechanism.

When `c_patient_class` is not included in your ESSENCE pull, the function
issues an informational message and skips `patient_class_change` detection.
All other mechanism types remain fully functional.

> **Why classify before removing duplicates?**
> Classification reveals whether your duplicates are dominated by retransmissions
> (suggesting `keep = "last"` is appropriate) or by `visit_date_change` events
> (which may warrant examining whether the date change affects your analysis
> window). It also identifies whether a specific facility is generating an
> unusual volume of a particular duplication type -- which can point to a
> feed-level configuration problem worth reporting to the facility or NSSP.

For row-level analysis, use `return_format = "tibble"` to join classifications
back to the original data:

```{r classify-tibble}
# Join mechanism type back to raw data for row-level inspection.
# classify_duplicates() returns cleaned (snake_case) column names,
# so clean essence_raw first to align join keys.
typed <- essence_raw |>
  janitor::clean_names() |>
  dplyr::left_join(
    classify_duplicates(essence_raw, return_format = "tibble"),
    by = c("hospital_name", "visit_id")
  )

dplyr::count(typed, dup_type)
```

## Removing Duplicates: `dedupe()`

With duplicates characterized, remove them using `dedupe()`. The function
retains exactly one row per `facility × visit_col` group.

```{r dedupe}
deduped <- dedupe(essence_raw, order_by = Arrived_Date_Time, keep = "last")

cat("Rows before: ", nrow(essence_raw), "\n")
cat("Rows after:  ", nrow(deduped), "\n")
cat("Rows removed:", nrow(essence_raw) - nrow(deduped), "\n")
```

```{r verify-zero}
# Confirm zero duplicate groups remain
summarize_duplicates(deduped)$overall
```

The `order_by` argument specifies which column to use for ranking rows within
each duplicate group. `Arrived_Date_Time` -- the timestamp when NSSP received
the record -- is the recommended ordering column for production surveillance
because the most recently received row is most likely to reflect updated
clinical information.

## Choosing a Deduplication Strategy

The appropriate strategy depends on your surveillance context:

| Scenario | Recommended strategy |
|----------|---------------------|
| Rolling surveillance (production monitoring) | `order_by = Arrived_Date_Time, keep = "last"` |
| Fixed-window retrospective analysis | `order_by = Arrived_Date_Time, keep = "first"` |
| Field completeness matters (e.g., maximizing diagnosis code coverage) | `keep = "most_complete"` |
| Unsure -- inspect before deciding | Call `classify_duplicates()` first |

**`keep = "last"`** retains the row with the highest value of `order_by` -- the
most recently transmitted record. Appropriate when the latest transmission is
most current and complete, which is true for most retransmission scenarios.

**`keep = "first"`** retains the row with the lowest value of `order_by` --
the original transmission. Appropriate for fixed-window analyses where you want
to capture the visit as it was first reported, without post-hoc field updates.

**`keep = "most_complete"`** retains the row with the fewest `NA` values across
all columns. Appropriate when no single transmission is reliably more current
than another and field completeness is the primary concern -- for example, when
discharge diagnosis codes are populated inconsistently across retransmissions.

When the `order_by` column itself contains ties (multiple rows with the same
`Arrived_Date_Time`), `dedupe()` retains the first occurrence among tied rows
and issues an informational message. To avoid tie-breaking ambiguity, use
`C_BioSense_ID` as a secondary sort or combine `keep = "most_complete"` with
an explicit `order_by`.

## Next Steps

- `vignette("encounter-linkage")`: After deduplication, understand how
  `link_encounters()` handles the structural exclusion of direct admissions
  from `HasBeenE = 1` filtered queries.

- `vignette("getting-started")`: The full recommended pipeline, combining
  deduplication with care setting filtering and geographic attribution.

- `?dedupe`, `?summarize_duplicates`, `?classify_duplicates`: Full parameter
  documentation and additional examples.
