---
title: "Linking ED and Inpatient Records for Accurate Burden Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linking ED and Inpatient Records for Accurate Burden Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(sysPrep)
library(dplyr)
```

## The Care Pathway Artifact Problem

ESSENCE has no mechanism to detect when a record with `HasBeenE = 0` and
`HasBeenAdmitted = 1` is actually the continuation of a `HasBeenE = 1`
record submitted moments earlier at the same facility for the same visit,
rather than a genuinely independent admission. A single row showing
`HasBeenAdmitted = 1` and `HasBeenE = 0` (`C_Patient_Class = "I"`) can mean
either of two very different things:

- **A true direct admission.** The patient was referred from primary care,
  outpatient care, or urgent care -- or admitted based on a pre-arranged
  plan -- and genuinely never went through the ED at this facility. This is
  a real clinical pathway, not a data quality problem.
- **A mis-submitted continuity break.** The patient *was* triaged and
  treated in the ED first, but the facility's system closed that encounter
  as a discharge instead of tracking the ED-to-inpatient transition as one
  continuous record. The result is two records -- one correctly showing
  `HasBeenE = 1`, one showing `HasBeenAdmitted = 1` and `HasBeenE = 0` --
  that share the same `facility_col x visit_col` key but otherwise look, to
  ESSENCE, like two unrelated encounters. This is a data quality artifact,
  not a distinct clinical pathway.

Nothing in a single row distinguishes these two cases. The only way to tell
them apart is to check whether a matching ED record exists under the same
`facility_col x visit_col` key -- which is exactly what `link_encounters()`
does.

This has a direct, structural consequence for burden estimation. ESSENCE
has no built-in capacity to deduplicate a mis-submitted direct-admit record
against its originating ED visit. Running a `HasBeenE = 1` query and
combining it with an otherwise-identical `HasBeenE = 0` / `HasBeenAdmitted
= 1` query -- expecting the sum to be an accurate "severe healthcare visit"
census -- does not produce one: some proportion of the records in the
second query already exist as a separate ED visit in the first, and
combining them without deduplication double-counts that encounter. At the
same time, genuinely new direct admissions in the second query are exactly
what you're trying to capture. `link_encounters()` resolves both problems
at once: records sharing a `facility_col x visit_col` key are merged into
one true encounter, while direct-admit records with no matching ED record
are retained as their own distinct encounters.

In Kentucky's overdose surveillance data, some of the decline in the
`HasBeenE = 1` trend over time is genuinely explained by growth in the
`HasBeenE = 0` / `HasBeenAdmitted = 1` trend -- but not as much as naively
binning the two would suggest, since some proportion of that apparent
direct-admit growth is the same double-counting artifact described above,
not a new encounter. Reimbursement or clinical-protocol changes, improving
HL7/EHR data quality over time, or both together are plausible contributors
to a genuine pathway shift, but this can't be attributed to either cause
specifically. Practitioners unaware of this risk face it in two directions:
naively summing separately-queried `HasBeenE = 1` and direct-admit counts
overstates the true combined severe-visit census, while naively comparing
the two trends to explain a decline in one against growth in the other
overstates how much of that decline reflects a genuine pathway shift.

> **How common is this, and would I know if it were happening in my data?**
> This isn't a widely documented ESSENCE data quality issue. Using
> `HasBeenE = 1` and direct-admit data together for burden estimation isn't
> itself a widely formalized practice in ESSENCE -- in print or elsewhere --
> so the risk rarely surfaces. It came out of record-level review of
> Kentucky's data, not published guidance. Most users working from
> pre-aggregated `HasBeenE = 1` and `HasBeenAdmitted = 1` counts through
> timeSeries or tableBuilder have no way to detect it, since the
> double-counted encounters are indistinguishable from genuine direct
> admits at the aggregate level.

`link_encounters()` addresses this by combining an ED pull with a
separately queried inpatient pull, then constructing unified care episodes
that count each patient interaction once regardless of which pathway they
followed.

> **Could you address this by filtering to `HasBeenAdmitted = 1` instead?**
> A query filtered to `HasBeenAdmitted = 1` returns all admitted visits --
> including those with `HasBeenE = 1`, which are already present in the ED
> pull. Using only the inpatient pull would miss all ED visits that did not
> result in admission, which represent the majority of emergency encounters.
> `link_encounters()` resolves this by binding both pulls and marking the
> ED row as the index encounter for visits that appear in both.

## Patient Class Derivation

`link_encounters()` supports two methods for determining patient class from
ESSENCE data. The preferred method uses `C_Patient_Class_List`; the fallback
uses `HasBeen_` indicator flags.

### Preferred: `C_Patient_Class_List`

`C_Patient_Class_List` is an ESSENCE-computed field that contains an
alphabetic, deduplicated list of all `C_Patient_Class` values present across
all messages sharing the same ESSENCE ID. For a patient who was registered in
the ED (`E`) and subsequently admitted (`I`), the field contains `"EI"`.

The patient class codes follow the HL7 Patient Class standard as adopted by
the Public Health Information Network (PHIN). The full value set is
maintained by the
[CDC PHIN Vocabulary Access and Distribution System (PHIN VADS)](https://phinvads.cdc.gov/vads/ViewValueSet.action?id=564F8F8B-E1DE-E411-8970-0017A477041A),
whose preferred concept names are shown below.
`link_encounters()`'s own derived `patient_class` output values (`"ED"`,
`"Inpatient"`, `"Direct Admit"`, etc.) are shortened working labels, not
required to match these preferred names verbatim:

| Code | PHIN VADS preferred name |
|------|--------------------------|
| `D`  | Direct admit |
| `E`  | Emergency |
| `I`  | Inpatient |
| `V`  | Observation patient |
| `B`  | Obstetrics |
| `O`  | Outpatient |
| `P`  | Preadmit |
| `R`  | Recurring patient |

`C_Patient_Class_List` is strictly more informative than `HasBeen_` pivoting
for one important reason: it is computed across all messages for the visit
before deduplication, meaning the full care pathway history is preserved
regardless of which row `dedupe()` retained. A `HasBeen_` flag on a
deduplicated record reflects only the flags present on the retained row; if
`keep = "first"` was used and the first row happened to predate an inpatient
admission update, the admission flag may be absent even though the patient was
eventually admitted. `C_Patient_Class_List` does not have this limitation.

When `C_Patient_Class_List` is present in `ed_data`, `link_encounters()` uses
it automatically and informs the user. No `HasBeen_` fields are required.

### Fallback: `HasBeen_` flag pivot

When `C_Patient_Class_List` is not present, `link_encounters()` falls back to
pivoting the binary `HasBeen_` flag columns. The two most relevant flags for
encounter linkage are:

| `HasBeenE` | `HasBeenAdmitted` | Clinical interpretation |
|-----------|-------------------|------------------------|
| `1` | `0` | ED visit, discharged or left without being admitted |
| `1` | `1` | ED visit that escalated to inpatient admission |
| `0` | `1` | Direct admission -- no ED visit in this encounter |
| `0` | `0` | Not applicable (neither ED nor inpatient) |

A `HasBeenE = 1` filtered query returns rows in the first two patterns only.
The third pattern -- direct admissions -- requires a separate `HasBeenAdmitted
= 1` query to capture.

`HasBeenI` is an alternative inpatient flag available in some ESSENCE
installations. When both `HasBeenAdmitted` and `HasBeenI` are present,
`link_encounters()` uses `HasBeenAdmitted` preferentially because it is
discharge-disposition aware: it captures ED visits that resulted in inpatient
admission via discharge disposition codes, which `HasBeenI` may not flag in
all ESSENCE configurations. `essence_ed_raw` (introduced below) includes
both columns for its escalation visits, so this preference is visible as an
informational message the first time `link_encounters()` runs on it.

## Linking ED and Inpatient Pulls

`link_encounters()` always requires two separately queried ESSENCE pulls: an
ED pull (`HasBeenE = 1`) and an inpatient pull (`HasBeenAdmitted = 1` or
`HasBeenI = 1`). There is no single-pull mode. A `HasBeenE = 1` pull cannot
resolve the care pathway artifact problem on its own in either direction: a
genuine direct admission (`HasBeenE = 0`) is structurally absent from it by
construction, so there is nothing to link against; and an ED-to-inpatient
escalation that *is* visible within it already lives on one
already-deduplicated record (both `HasBeenE` and `HasBeenAdmitted` set),
which needs no linking, since there was never a second record to link
against. `link_encounters()` aborts if `inpatient_admission_data` is
omitted, rather than silently returning `ed_data` unchanged with cosmetic
metadata columns appended.

`sysPrep` ships two small synthetic datasets purpose-built for this
demonstration: `essence_ed_raw` (a `HasBeenE = 1` pull) and
`essence_inp_raw` (a separately queried `HasBeenAdmitted = 1` pull) --
distinct from the larger, general-purpose `essence_raw` used elsewhere in
this documentation, since `essence_raw` represents a single realistic ED
pull and was never meant to be split into two:

```{r glimpse-pulls}
nrow(essence_ed_raw)
nrow(essence_inp_raw)
```

Deduplicate each pull separately before linking:

```{r prepare}
ed_clean <- essence_ed_raw |>
  dedupe(order_by = Arrived_Date_Time, keep = "last") |>
  filter_care_setting(
    fix_facility_type_vector = c("Hillside FSED", "Downtown Emergency Services")
  )

inpatient_clean <- essence_inp_raw |>
  dedupe(order_by = Arrived_Date_Time, keep = "last")
```

> **Can I just row-bind the ED and admission pulls and run `dedupe()`
> on the result, instead of passing them separately?**
> No -- this risks silently losing visits. `dedupe()`'s
> `facility_col x visit_col` grouping does not distinguish an ED record
> from its corresponding direct-admit record; from `dedupe()`'s
> perspective they are simply two rows sharing a key. `keep = "last"` or
> `keep = "first"` will pick whichever row has the later or earlier
> `Arrived_Date_Time` and discard the other -- unpredictably decreasing
> either the ED visit count or the direct-admit count, depending on that
> ordering, rather than merging the two into one complete encounter.
> Always deduplicate each pull separately (as shown above) and pass both
> to `link_encounters()` as separate arguments; let `link_encounters()`,
> not `dedupe()`, perform the combination.

By default (`return_format = "collapsed"`), `link_encounters()` merges each
episode's rows into a single composite row rather than returning long-format
duplicates that must be filtered afterward. Rows with `HasBeenE = 1` in the
admission pull are automatically removed first (they are already present in
the ED pull); the remainder are labeled `patient_class = "Direct Admit"` and
combined with the ED pull's own rows. Within a merged episode, `HasBeen_`
flags are taken as the max across the episode's rows, and fields named in
`merge_fields` (`CCDD`, `CCDDParsed`, `CCDDCategory_flat`, `C_Death`,
`Discharge_Disposition`, and `DispositionCategory` by default) are
reconciled using a strategy suited to that field, so information recorded
only on the admission row is not silently discarded when the ED row is kept
as the surviving primary row.

```{r episodes}
episodes <- link_encounters(ed_clean, inpatient_clean)

dplyr::glimpse(episodes)
```

Every row above shows real `0`/`1` values for `has_been_e`/`has_been_admitted`
-- never `NA` -- regardless of whether it came from a single pre-merge row or
a merged pair: deriving `patient_class` from `HasBeen_` flags (the fallback
used here, since this synthetic data has no `C_Patient_Class_List`) pivots
those flag columns to build `patient_class`, but `link_encounters()` retains
the true original values underneath so they survive into the output
unchanged.

`nrow(episodes)` is already the true unduplicated encounter count -- smaller
than the naive combined denominator, since some admission-pull rows merged
into an existing ED episode instead of adding a new one:

```{r episode-count}
nrow(ed_clean) + nrow(inpatient_clean) # naive combined denominator
nrow(episodes)                         # true unduplicated encounter count
```

Four episode metadata columns are added to every row, regardless of
`return_format`:

**`.episode_id`** -- A character key combining facility and visit identifier,
shared across all rows belonging to the same encounter. Used internally to
group rows into episodes and available for joining or filtering downstream.

**`.patient_class_sequence`** -- A collapsed string of the patient classes
present in the episode, in the order they actually occurred (using
`C_Visit_Date_Time`, `Date`+`Time`, or `C_Patient_Class_MDT_Updates` when
available -- see `?link_encounters`). The ED and admission records are
genuinely separate rows with their own timestamps, so this can reflect a
true reversal -- e.g. `"Direct Admit->ED"` if an admission record's
timestamp somehow precedes the ED record's, versus the expected
`"ED->Direct Admit"` for a normal continuity-break artifact (ED visit,
then mis-submitted admission continuation). An episode built from a single
row has `.patient_class_sequence` equal to that row's own class (e.g.
`"ED"`, or `"Direct Admit"` for a genuine direct admission with no ED
counterpart). This column enables filtering to specific episode types
(e.g., `filter(.patient_class_sequence == "ED->Direct Admit")` to isolate
the continuity-break cases) even though the underlying rows have already
been merged.

**`.episode_n_rows`** -- The number of original rows the episode was built
from before merging. Episodes with no linkage have `.episode_n_rows = 1`;
episodes merged from an ED row and an admission row have
`.episode_n_rows = 2`.

**`.index_encounter`** -- In collapsed output (the default), always `TRUE`,
since every row already represents exactly one merged encounter; the column
is retained for schema consistency with the diagnostic long-format output
described below, not because a filtering step is required.

```{r burden-pathways}
# Distribution of care pathways -- already one row per encounter
episodes |>
  dplyr::count(.patient_class_sequence, sort = TRUE)
```

`"ED"` visits never touched the admission pull at all. `"Admitted->ED"`
visits escalated to inpatient care within the ED pull itself (`HasBeenE = 1`
and `HasBeenAdmitted = 1` on one already-deduplicated record) -- these need
no cross-pull linking, since there was only ever one record. `"Direct
Admit"` visits are genuine admissions with no preceding ED encounter at
all, structurally invisible to a `HasBeenE = 1` query and only captured
because a second pull was queried. `"ED->Direct Admit"` visits are the
continuity-break artifact itself: an ED visit and its mis-submitted
direct-admit continuation, reported under two different `C_BioSense_ID`
values sharing one `HospitalName x Visit_ID`, correctly merged into a
single encounter instead of being counted twice.

```{r direct-admits}
dplyr::filter(episodes, patient_class == "Direct Admit") |>
  dplyr::select(hospital_name, visit_id, .patient_class_sequence)
```

### Inspecting the Raw Linkage Mechanism: `return_format = "long"`

Set `return_format = "long"` to see the pre-merge rows directly -- for
example, to audit which fields differed between the ED and admission records
before they were reconciled, or to understand the linkage mechanism itself.
No merge is applied in this mode.

```{r long-format}
episodes_long <- link_encounters(
  ed_clean, inpatient_clean, return_format = "long"
)

dplyr::count(episodes_long, patient_class)
```

In long format, `.index_encounter` marks the row that survives a "one row per
episode" filter -- `TRUE` on the ED row when both ED and admission rows are
present, or on the single row for ED-only or Direct Admit-only episodes:

```{r long-index}
episodes_long |>
  dplyr::filter(.index_encounter) |>
  dplyr::count(.patient_class_sequence, sort = TRUE)
```

This matches the collapsed-format encounter count above, since the same rows
are merged into (or, in long format, selected as) the encounter of record.

## Burden Estimation from Linked Data

With the default collapsed output, unduplicated burden estimation is just a
count -- no `.index_encounter` filter is needed, since every row already
represents one merged encounter:

```{r burden}
burden <- episodes |>
  dplyr::count(.patient_class_sequence, sort = TRUE)

burden
```

This yields one count per clinical encounter. Visits that generated both an
ED signal and an admission signal are merged into a single row (using the ED
row as primary) and counted once. Visits that generated only an ED signal are
counted once. Direct admissions add their own counts without inflating the
ED rows.

The `.patient_class_sequence` breakdown is informative for understanding the
severity composition of the case count:

- `"ED"` -- Emergency department visit without inpatient admission
- `"Admitted->ED"` -- Emergency department visit that escalated to inpatient
  admission within the ED pull alone (merged into a single row, primary
  values from the ED row)
- `"Direct Admit"` -- Genuine inpatient admission with no preceding ED
  encounter at all, only visible because a second pull was queried
- `"ED->Direct Admit"` -- The continuity-break artifact: an ED visit and
  its mis-submitted direct-admit continuation, reported as two records and
  correctly merged into one encounter

For most ESSENCE-based incidence analyses, the sum across all
`.patient_class_sequence` categories in the (already unduplicated) collapsed
output is the appropriate unduplicated encounter count.

## Connection to Cluster Detection

If using linked encounter data as input to spatial cluster detection
algorithms (e.g., SaTScan, the `gsClusterDetect` package, or custom
Kulldorff scan implementations), the default collapsed output is already one
row per encounter, so no additional deduplication filter is needed before
aggregating counts to geographic units:

```{r cluster-prep, eval = FALSE}
# Prepare for spatial analysis: already one row per encounter
episodes |>
  assign_treating_geography() |>
  dplyr::count(region_hybrid)
```

If you instead requested `return_format = "long"` for diagnostic purposes,
filter to `.index_encounter == TRUE` first -- otherwise multi-class episodes
would be counted twice, once per pre-merge row, inflating cluster statistics.
