---
title: "Linking ED and Inpatient Records for Accurate Burden Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linking ED and Inpatient Records for Accurate Burden Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(sysPrep)
library(dplyr)
```

## The Care Pathway Artifact Problem

Standard ESSENCE surveillance queries filtered to `HasBeenE = 1` return visits
where a patient received emergency department care. This is the appropriate
filter for ED-based syndromic surveillance -- but it has a structural
consequence that practitioners rarely account for: it cannot return visits
where `HasBeenE = 0` and `HasBeenAdmitted = 1`. These are **direct
admissions** -- patients admitted to inpatient units without ED triage.

Direct admissions are not a data quality problem. They represent a real
clinical pathway in which a patient, typically through a physician referral or
a pre-arranged admission, bypasses the emergency department entirely and is
admitted directly to a medical, surgical, or telemetry unit. From a public
health surveillance perspective, the question is whether changes in the
proportion of patients following this pathway can be misidentified as changes
in disease burden.

They can. During opioid overdose surveillance in Kentucky, an apparent
near-linear decline in emergency department opioid overdose visit volume was
observed over a multi-year period. Detailed investigation revealed that a
substantial portion of this apparent trend was attributable to a systematic
shift in hospital admission practices: facilities in certain markets had begun
directly admitting opioid overdose patients -- routing them to telemetry units
without ED triage -- in response to changes in reimbursement structures and
clinical protocols. From a `HasBeenE = 1` query perspective, these patients
were invisible. The observed "decline" was, in part, a care pathway artifact
rather than a true reduction in overdose incidence.

The policy implications are serious. An apparent decline in overdose burden
might reduce intervention prioritization, divert resources away from
prevention programs, and lead surveillance reports to understate the severity
of an ongoing crisis -- at precisely the time the actual burden may be stable
or increasing.

`link_encounters()` addresses this by combining an ED pull with a separately
queried inpatient pull, then constructing unified care episodes that count
each patient interaction once regardless of which pathway they followed.

> **Could you address this by filtering to `HasBeenAdmitted = 1` instead?**
> A query filtered to `HasBeenAdmitted = 1` returns all admitted visits --
> including those with `HasBeenE = 1`, which are already present in the ED
> pull. Using only the inpatient pull would miss all ED visits that did not
> result in admission, which represent the majority of emergency encounters.
> `link_encounters()` resolves this by binding both pulls and marking the
> ED row as the index encounter for visits that appear in both.

## Patient Class Derivation

`link_encounters()` supports two methods for determining patient class from
ESSENCE data. The preferred method uses `C_Patient_Class_List`; the fallback
uses `HasBeen_` indicator flags.

### Preferred: `C_Patient_Class_List`

`C_Patient_Class_List` is an ESSENCE-computed field that contains an
alphabetic, deduplicated list of all `C_Patient_Class` values present across
all messages sharing the same ESSENCE ID. For a patient who was registered in
the ED (`E`) and subsequently admitted (`I`), the field contains `"EI"`.

The patient class codes follow the HL7 Patient Class standard as adopted by
the Public Health Information Network (PHIN). The full value set is maintained
by the [CDC PHIN Vocabulary Access and Distribution System (PHIN VADS)](https://phinvads.cdc.gov/ViewValueSet.action?id=564F8F8B-E1DE-E411-8970-0017A477041A):

| Code | Patient class |
|------|--------------|
| `E`  | ED |
| `I`  | Inpatient |
| `D`  | Direct Admit |
| `V`  | Observation |
| `O`  | Outpatient |
| `B`  | Obstetrics |
| `P`  | Preadmit |
| `R`  | Recurring |

`C_Patient_Class_List` is strictly more informative than `HasBeen_` pivoting
for one important reason: it is computed across all messages for the visit
before deduplication, meaning the full care pathway history is preserved
regardless of which row `dedupe()` retained. A `HasBeen_` flag on a
deduplicated record reflects only the flags present on the retained row; if
`keep = "first"` was used and the first row happened to predate an inpatient
admission update, the admission flag may be absent even though the patient was
eventually admitted. `C_Patient_Class_List` does not have this limitation.

When `C_Patient_Class_List` is present in `ed_data`, `link_encounters()` uses
it automatically and informs the user. No `HasBeen_` fields are required.

### Fallback: `HasBeen_` flag pivot

When `C_Patient_Class_List` is not present, `link_encounters()` falls back to
pivoting the binary `HasBeen_` flag columns. The two most relevant flags for
encounter linkage are:

| `HasBeenE` | `HasBeenAdmitted` | Clinical interpretation |
|-----------|-------------------|------------------------|
| `1` | `0` | ED visit, discharged or left without being admitted |
| `1` | `1` | ED visit that escalated to inpatient admission |
| `0` | `1` | Direct admission -- no ED visit in this encounter |
| `0` | `0` | Not applicable (neither ED nor inpatient) |

A `HasBeenE = 1` filtered query returns rows in the first two patterns only.
The third pattern -- direct admissions -- requires a separate `HasBeenAdmitted
= 1` query to capture.

`HasBeenI` is an alternative inpatient flag available in some ESSENCE
installations. When both `HasBeenAdmitted` and `HasBeenI` are present,
`link_encounters()` uses `HasBeenAdmitted` preferentially because it is
discharge-disposition aware: it captures ED visits that resulted in inpatient
admission via discharge disposition codes, which `HasBeenI` may not flag in
all ESSENCE configurations.

## Single-Pull Linkage

By default (`return_format = "collapsed"`), `link_encounters()` merges each
episode's rows into a single composite row rather than returning long-format
duplicates that must be filtered afterward. A visit with `HasBeenE = 1` and
`HasBeenAdmitted = 1` is still detected internally as two rows -- one
representing the ED contact, one representing the inpatient escalation -- but
those rows are reconciled into one merged row: `HasBeen_` flags are taken as
the max across the episode's rows, and fields named in `merge_fields`
(`CCDD`, `CCDDParsed`, `CCDDCategory_flat`, `C_Death`, `Discharge_Disposition`,
and `DispositionCategory` by default) are reconciled using a strategy suited
to that field, so information recorded only on the inpatient row is not
silently discarded when the ED row is kept as the surviving primary row.

```{r prepare}
# Deduplicate and filter before linking
ed_clean <- essence_raw |>
  dedupe(order_by = Arrived_Date_Time, keep = "last") |>
  filter_care_setting(
    fix_facility_type_vector = c("Hillside FSED", "Downtown Emergency Services")
  )
```

```{r single-pull}
episodes <- link_encounters(ed_clean)

dplyr::glimpse(episodes)
```

`nrow(episodes)` is already the unduplicated encounter count -- no additional
filtering step is required with the default collapsed output:

```{r episode-count}
nrow(episodes)
```

Four episode metadata columns are added to every row, regardless of
`return_format`:

**`.episode_id`** -- A character key combining facility and visit identifier,
shared across all rows belonging to the same encounter. Used internally to
group rows into episodes and available for joining or filtering downstream.

**`.patient_class_sequence`** -- A collapsed string of the patient classes
present in the episode, in the order they actually occurred (using
`C_Visit_Date_Time`, `Date`+`Time`, or `C_Patient_Class_MDT_Updates` when
available -- see `?link_encounters`). A single-pull ED record with both
`HasBeenE = 1` and `HasBeenAdmitted = 1` set has no way to distinguish when
each flag was assigned, so its two classes tie and fall back to alphabetical
order: `.patient_class_sequence = "Admitted->ED"`. Two-pull linkage, where
the ED and direct-admit records are genuinely separate rows with their own
timestamps, can instead reflect a true reversal -- e.g.
`"Direct Admit->ED"` when the direct-admit record's timestamp precedes the
ED record's. An episode built from a single ED row has
`.patient_class_sequence = "ED"`. This column enables filtering to specific
episode types (e.g., `filter(.patient_class_sequence == "Admitted->ED")` to
isolate escalating visits) even though the underlying rows have already been
merged.

**`.episode_n_rows`** -- The number of original rows the episode was built
from before merging. Episodes with no escalation have `.episode_n_rows = 1`;
ED + Admitted episodes collapsed from two original rows have
`.episode_n_rows = 2`.

**`.index_encounter`** -- In collapsed output (the default), always `TRUE`,
since every row already represents exactly one merged encounter; the column
is retained for schema consistency with the diagnostic long-format output
described below, not because a filtering step is required.

```{r burden-single}
# Distribution of care pathways -- already one row per encounter
episodes |>
  dplyr::count(.patient_class_sequence, sort = TRUE)
```

### Inspecting the Raw Linkage Mechanism: `return_format = "long"`

Set `return_format = "long"` to see the pre-merge rows directly -- for
example, to audit which fields differed between the ED and inpatient records
before they were reconciled, or to understand the linkage mechanism itself.
No merge is applied in this mode.

```{r long-format}
episodes_long <- link_encounters(ed_clean, return_format = "long")

dplyr::count(episodes_long, patient_class)
```

In long format, `.index_encounter` marks the row that survives a "one row per
episode" filter -- `TRUE` on the ED row when both ED and Admitted rows are
present, or on the single row for ED-only or Direct Admit-only episodes:

```{r long-index}
episodes_long |>
  dplyr::filter(.index_encounter) |>
  dplyr::count(.patient_class_sequence, sort = TRUE)
```

This matches the collapsed-format encounter count above, since the same rows
are merged into (or, in long format, selected as) the encounter of record.

## Two-Pull Linkage

To capture direct admissions, run a second ESSENCE query filtered to
`HasBeenAdmitted = 1` (or `HasBeenI = 1`), deduplicate it separately, and
supply it as `inpatient_admission_data`. Rows with `HasBeenE = 1` in the
inpatient pull are automatically removed to prevent double-counting -- these
visits are already present in the ED pull.

```{r two-pull, eval = FALSE}
# Query ESSENCE for HasBeenAdmitted = 1 separately, then:
inpatient_clean <- essence_inpatient |>
  dedupe(order_by = Arrived_Date_Time, keep = "last")

episodes_full <- link_encounters(ed_clean, inpatient_clean)

# Direct admissions are now visible
dplyr::filter(episodes_full, patient_class == "Direct Admit")
```

When `inpatient_admission_data` is supplied, `link_encounters()` removes rows
with `HasBeenE = 1` from that pull (informing you of how many were dropped)
and labels the remaining rows `patient_class = "Direct Admit"` before
merging. A direct-admit visit that shares no facility x visit_id with an ED
row has no rows to merge with, so it survives collapsing as its own
single-row episode with `.patient_class_sequence = "Direct Admit"`.

## Burden Estimation from Linked Data

With the default collapsed output, unduplicated burden estimation is just a
count -- no `.index_encounter` filter is needed, since every row already
represents one merged encounter:

```{r burden}
burden <- episodes |>
  dplyr::count(.patient_class_sequence, sort = TRUE)

burden
```

This yields one count per clinical encounter. Visits that generated both an
ED signal and an Admitted signal are merged into a single row (using the ED
row as primary) and counted once. Visits that generated only an ED signal are
counted once. With two-pull linkage, direct admissions add their own counts
without inflating the ED rows.

The `.patient_class_sequence` breakdown is informative for understanding the
severity composition of the case count:

- `"ED"` -- Emergency department visit without inpatient admission
- `"Admitted->ED"` -- Emergency department visit that escalated to inpatient
  admission (merged into a single row, primary values from the ED row)
- `"Direct Admit"` -- Inpatient admission without ED encounter (only visible
  with two-pull linkage)

For most ESSENCE-based incidence analyses, the sum across all
`.patient_class_sequence` categories in the (already unduplicated) collapsed
output is the appropriate unduplicated encounter count.

## Connection to Cluster Detection

If using linked encounter data as input to spatial cluster detection
algorithms (e.g., SaTScan, the `gsClusterDetect` package, or custom
Kulldorff scan implementations), the default collapsed output is already one
row per encounter, so no additional deduplication filter is needed before
aggregating counts to geographic units:

```{r cluster-prep, eval = FALSE}
# Prepare for spatial analysis: already one row per encounter
episodes |>
  assign_treating_geography() |>
  dplyr::count(region)
```

If you instead requested `return_format = "long"` for diagnostic purposes,
filter to `.index_encounter == TRUE` first -- otherwise multi-class episodes
would be counted twice, once per pre-merge row, inflating cluster statistics.
