---
title: "Linking ED and Inpatient Records for Accurate Burden Estimation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Linking ED and Inpatient Records for Accurate Burden Estimation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(sysPrep)
library(dplyr)
```

## The Care Pathway Artifact Problem

Standard ESSENCE surveillance queries filtered to `HasBeenE = 1` return visits
where a patient received emergency department care. This is the appropriate
filter for ED-based syndromic surveillance -- but it has a structural
consequence that practitioners rarely account for: it cannot return visits
where `HasBeenE = 0` and `HasBeenAdmitted = 1`. These are **direct
admissions** -- patients admitted to inpatient units without ED triage.

Direct admissions are not a data quality problem. They represent a real
clinical pathway in which a patient, typically through a physician referral or
a pre-arranged admission, bypasses the emergency department entirely and is
admitted directly to a medical, surgical, or telemetry unit. From a public
health surveillance perspective, the question is whether changes in the
proportion of patients following this pathway can be misidentified as changes
in disease burden.

They can. During opioid overdose surveillance in Kentucky, an apparent
near-linear decline in emergency department opioid overdose visit volume was
observed over a multi-year period. Detailed investigation revealed that a
substantial portion of this apparent trend was attributable to a systematic
shift in hospital admission practices: facilities in certain markets had begun
directly admitting opioid overdose patients -- routing them to telemetry units
without ED triage -- in response to changes in reimbursement structures and
clinical protocols. From a `HasBeenE = 1` query perspective, these patients
were invisible. The observed "decline" was, in part, a care pathway artifact
rather than a true reduction in overdose incidence.

The policy implications are serious. An apparent decline in overdose burden
might reduce intervention prioritization, divert resources away from
prevention programs, and lead surveillance reports to understate the severity
of an ongoing crisis -- at precisely the time the actual burden may be stable
or increasing.

`link_encounters()` addresses this by combining an ED pull with a separately
queried inpatient pull, then constructing unified care episodes that count
each patient interaction once regardless of which pathway they followed.

> **Could you address this by filtering to `HasBeenAdmitted = 1` instead?**
> A query filtered to `HasBeenAdmitted = 1` returns all admitted visits --
> including those with `HasBeenE = 1`, which are already present in the ED
> pull. Using only the inpatient pull would miss all ED visits that did not
> result in admission, which represent the majority of emergency encounters.
> `link_encounters()` resolves this by binding both pulls and marking the
> ED row as the index encounter for visits that appear in both.

## Patient Class Derivation

`link_encounters()` supports two methods for determining patient class from
ESSENCE data. The preferred method uses `C_Patient_Class_List`; the fallback
uses `HasBeen_` indicator flags.

### Preferred: `C_Patient_Class_List`

`C_Patient_Class_List` is an ESSENCE-computed field that contains an
alphabetic, deduplicated list of all `C_Patient_Class` values present across
all messages sharing the same ESSENCE ID. For a patient who was registered in
the ED (`E`) and subsequently admitted (`I`), the field contains `"EI"`.

The patient class codes follow the HL7 Patient Class standard as adopted by
the Public Health Information Network (PHIN). The full value set is maintained
by the [CDC PHIN Vocabulary Access and Distribution System (PHIN VADS)](https://phinvads.cdc.gov/ViewValueSet.action?id=564F8F8B-E1DE-E411-8970-0017A477041A):

| Code | Patient class |
|------|--------------|
| `E`  | ED |
| `I`  | Inpatient |
| `D`  | Direct Admit |
| `V`  | Observation |
| `O`  | Outpatient |
| `B`  | Obstetrics |
| `P`  | Preadmit |
| `R`  | Recurring |

`C_Patient_Class_List` is strictly more informative than `HasBeen_` pivoting
for one important reason: it is computed across all messages for the visit
before deduplication, meaning the full care pathway history is preserved
regardless of which row `dedupe()` retained. A `HasBeen_` flag on a
deduplicated record reflects only the flags present on the retained row; if
`keep = "first"` was used and the first row happened to predate an inpatient
admission update, the admission flag may be absent even though the patient was
eventually admitted. `C_Patient_Class_List` does not have this limitation.

When `C_Patient_Class_List` is present in `ed_data`, `link_encounters()` uses
it automatically and informs the user. No `HasBeen_` fields are required.

### Fallback: `HasBeen_` flag pivot

When `C_Patient_Class_List` is not present, `link_encounters()` falls back to
pivoting the binary `HasBeen_` flag columns. The two most relevant flags for
encounter linkage are:

| `HasBeenE` | `HasBeenAdmitted` | Clinical interpretation |
|-----------|-------------------|------------------------|
| `1` | `0` | ED visit, discharged or left without being admitted |
| `1` | `1` | ED visit that escalated to inpatient admission |
| `0` | `1` | Direct admission -- no ED visit in this encounter |
| `0` | `0` | Not applicable (neither ED nor inpatient) |

A `HasBeenE = 1` filtered query returns rows in the first two patterns only.
The third pattern -- direct admissions -- requires a separate `HasBeenAdmitted
= 1` query to capture.

`HasBeenI` is an alternative inpatient flag available in some ESSENCE
installations. When both `HasBeenAdmitted` and `HasBeenI` are present,
`link_encounters()` uses `HasBeenAdmitted` preferentially because it is
discharge-disposition aware: it captures ED visits that resulted in inpatient
admission via discharge disposition codes, which `HasBeenI` may not flag in
all ESSENCE configurations.

## Single-Pull Linkage

Even without a separate inpatient pull, `link_encounters()` adds analytical
value by pivoting the `HasBeen_` flag columns to long format. A visit with
`HasBeenE = 1` and `HasBeenAdmitted = 1` produces two rows -- one labeled
`"ED"` and one labeled `"Admitted"` -- along with episode metadata that
supports burden counting.

```{r prepare}
# Deduplicate and filter before linking
ed_clean <- essence_raw |>
  dedupe(order_by = Arrived_Date_Time, keep = "last") |>
  filter_care_setting(
    fix_facility_type_vector = c("Hillside FSED", "Downtown Emergency Services")
  )
```

```{r single-pull}
episodes <- link_encounters(ed_clean)

dplyr::glimpse(episodes)
```

```{r patient-class-counts}
# Distribution of patient class rows
dplyr::count(episodes, patient_class)
```

Four episode metadata columns are added to every row:

**`.episode_id`** -- A character key combining facility and visit identifier,
shared across all rows belonging to the same encounter. Used internally to
group rows into episodes and available for joining or filtering downstream.

**`.patient_class_sequence`** -- A sorted, collapsed string of all patient
classes in the episode. A visit with both ED and Admitted rows has
`.patient_class_sequence = "Admitted->ED"`. A visit with only an ED row has
`.patient_class_sequence = "ED"`. This column enables filtering to specific
episode types (e.g., `filter(.patient_class_sequence == "Admitted->ED")` to
isolate escalating visits).

**`.episode_n_rows`** -- The number of rows in the episode. Single-class
episodes have `.episode_n_rows = 1`; ED + Admitted episodes have
`.episode_n_rows = 2`.

**`.index_encounter`** -- `TRUE` on the ED row when both ED and Admitted rows
are present; `TRUE` on the single row for ED-only or Direct Admit-only
episodes. Filtering to `.index_encounter == TRUE` yields exactly one row per
clinical encounter -- the appropriate denominator for unduplicated burden counts.

```{r burden-single}
# Unduplicated episode count (one row per encounter)
episodes |>
  dplyr::filter(.index_encounter) |>
  dplyr::count(.patient_class_sequence, sort = TRUE)
```

## Two-Pull Linkage

To capture direct admissions, run a second ESSENCE query filtered to
`HasBeenAdmitted = 1` (or `HasBeenI = 1`), deduplicate it separately, and
supply it as `inpatient_admission_data`. Rows with `HasBeenE = 1` in the
inpatient pull are automatically removed to prevent double-counting -- these
visits are already present in the ED pull.

```{r two-pull, eval = FALSE}
# Query ESSENCE for HasBeenAdmitted = 1 separately, then:
inpatient_clean <- essence_inpatient |>
  dedupe(order_by = Arrived_Date_Time, keep = "last")

episodes_full <- link_encounters(ed_clean, inpatient_clean)

# Direct admissions are now visible
dplyr::filter(episodes_full, patient_class == "Direct Admit")
```

When `inpatient_admission_data` is supplied, `link_encounters()` removes rows
with `HasBeenE = 1` from that pull (informing you of how many were dropped)
and labels the remaining rows as `patient_class = "Direct Admit"`. These rows
are then bound to the long-format ED data and receive `.index_encounter = TRUE`
because they represent an encounter with no corresponding ED row.

## Burden Estimation from Linked Data

The recommended approach for unduplicated burden estimation is to filter to
`.index_encounter == TRUE` and count:

```{r burden}
burden <- episodes |>
  dplyr::filter(.index_encounter) |>
  dplyr::count(.patient_class_sequence, sort = TRUE)

burden
```

This yields one count per clinical encounter. Visits that generated both ED
and Admitted rows are counted once (from the ED row, which is the index
encounter for that episode). Visits that generated only an ED row are counted
once. With two-pull linkage, direct admissions add their own counts without
inflating the ED rows.

The `.patient_class_sequence` breakdown is informative for understanding the
severity composition of the case count:

- `"ED"` -- Emergency department visit without inpatient admission
- `"Admitted->ED"` -- Emergency department visit that escalated to inpatient
  admission (counted from the ED row)
- `"Direct Admit"` -- Inpatient admission without ED encounter (only visible
  with two-pull linkage)

For most ESSENCE-based incidence analyses, the sum across all `.patient_class_sequence`
categories at `.index_encounter == TRUE` is the appropriate unduplicated
encounter count.

## Connection to Cluster Detection

If using linked encounter data as input to spatial cluster detection algorithms
(e.g., SaTScan, the `gsClusterDetect` package, or custom Kulldorff scan
implementations), filter to `.index_encounter == TRUE` before aggregating
counts to geographic units. Failing to do so would count multi-class episodes
twice -- once for the ED row and once for the Admitted row -- inflating cluster
statistics.

```{r cluster-prep, eval = FALSE}
# Prepare for spatial analysis: one row per encounter, with geographic field
episodes |>
  dplyr::filter(.index_encounter) |>
  assign_treating_geography() |>
  dplyr::count(region)
```
