---
title: "Geographic Attribution for Out-of-State and Unknown Residence Visits"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Geographic Attribution for Out-of-State and Unknown Residence Visits}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(sysPrep)
library(dplyr)
```

## Why Out-of-State Visits Exist

ESSENCE surveillance is facility-based: every patient treated at a
participating facility is included in the data, regardless of where that
patient lives. A facility near a state border or serving a metropolitan area
that spans multiple states will routinely treat patients whose address of
record is in a neighboring state. These patients generate valid, important
clinical encounters: they may have overdosed at a Kentucky location even if
their home is in Indiana or West Virginia.

Two additional sources of non-Kentucky geography appear in ESSENCE data:

**`OTHER_REGION`** records represent patients for whom a residential address is
unavailable to the treating facility. This includes unhoused patients,
patients who cannot communicate their address, and patients who decline to
provide one. The `OTHER_REGION` value is not a geographic unit; it is a
placeholder indicating that no residential geography can be assigned.

**`{SITE}_UNKNOWN`** (e.g. `"KY_UNKNOWN"`) records represent visits at a
known-site facility that did not transmit patient address fields. Unlike
`OTHER_REGION`, this value carries the site prefix; it is ESSENCE's
placeholder for "the system knows this residence is missing," as distinct
from a residence that was received but does not map to any region. In
practice, `{SITE}_UNKNOWN` is rare relative to `OTHER_REGION`: observed
volumes in production Kentucky surveillance run in the single digits
against tens of thousands of `OTHER_REGION` records over the same period,
but because it carries the site prefix, it was silently misclassified as
in-state and left unreassigned prior to this being caught. A bare `NA` in
`Region` is treated as unknown residence too, though in practice
ESSENCE typically fills in one of the placeholder values above rather than
leaving the field empty.

## The Cost of Discarding These Visits

The common response to out-of-state and unknown-residence records is to filter
them out before analysis. This is understandable: residential geography is
the expected denominator for incidence-based rates, but it introduces
systematic bias. And it rarely requires an explicit `filter()` call to happen:
any downstream step scoped to your jurisdiction's regions, joining to a
shapefile of in-state counties for a map, grouping by `Region` against a
fixed list of in-state values for a summary table, silently drops rows
whose `Region` doesn't match, with no filter statement in the code to flag
the decision for review.

**Burden understatement.** Facilities that treat a high volume of out-of-state
or transient patients see their case counts reduced proportionally. A border
facility that treats patients from three states appears to have a much lower
burden than its actual clinical workload warrants.

**Spatial bias.** The exclusion rate is not uniform across facilities.
Facilities in rural border communities and urban trauma centers serving mobile
populations will have much higher exclusion rates than facilities in
geographically isolated inland areas. When these exclusions are not accounted
for, spatial comparisons across facilities are measuring a mix of true
incidence and differential data completeness.

**Secular trend distortion.** If the composition of a facility's patient
population shifts over time, more transient patients, more border traffic, or
a policy change affecting unhoused populations, the exclusion of
`OTHER_REGION` records can create apparent trends in what is actually a stable
underlying burden.

This is also why `sysPrep` treats the fix as attribution rather than
exclusion-avoidance alone: EMS-based systems (e.g., ODMAP, Biospatial) report
incidence at the location where care was rendered, not the patient's
jurisdiction of residence. Assigning treating facility geography brings
ESSENCE data into that same incidence-based frame, which matters for applied,
rapid-response surveillance where the operative question is where a case was
treated, not which jurisdiction is responsible for the follow-up.

`sysPrep` provides two functions for retaining these visits by attributing them
to the geography of the treating facility. The choice of which function to use
depends on the research question.

## What `Region` Actually Represents

> **Is `Region` an authoritative county boundary?**
> No. `Region` is ESSENCE's standardized construct for enabling consistent
> sub-state reporting across data sources: a maintained, many-to-one
> zip-code-to-region lookup table, not a live geocode computed from each
> record. By default, a zip code is assigned to a region using its
> geographic centroid, but individual assignments can be, and routinely
> are, overridden by site or state administrators to better reflect where
> the bulk of a zip code's population actually lives. The same logic
> applies to `HospitalRegion`: assigned from the facility's zip centroid by
> default, but overridable to the facility's listed county. Because of this
> approximation, results reported by `Region` should not be construed as
> the authoritative count for that county; they are the best available
> standardized proxy, and the mapping in effect today is not necessarily
> the mapping that applied when a record was first received.

This matters for `sysPrep` because both geography functions operate
entirely on `Region`/`HospitalRegion` as received; they reassign which
existing value applies to a row, they do not independently geocode or
validate the underlying zip-to-county mapping.

It also explains *why* two different functions exist at all, rather than
one that always attributes to "the" geography. ESSENCE itself distinguishes
between **patient location** and **facility location** data sources:

- In a **patient location** pull (e.g., `va_er`), the display criterion is
  the patient's residence: the data show who, from a given jurisdiction,
  presented for care anywhere.
- In a **facility location** pull (e.g., `va_hosp`), the display criterion
  is the treating facility: the data show everyone who presented at an
  in-jurisdiction facility, combining residents and visitors alike.

`essence_raw` and `essence_clean` model a patient location pull, so
`Region` reflects patient residence out of the box.
`assign_treating_geography()` preserves that patient-location semantics for
the in-state majority and only substitutes facility geography as a
fallback. `assign_facility_geography()` instead converts the dataset to
facility-location semantics uniformly, useful when your source pull, or
your research question, is facility-location in nature to begin with.

## The `Region` Field Format

ESSENCE encodes geography in the format `{SITE}_{REGION}`, where `SITE` is
the NSSP Site Short Name (e.g., `"KY"`) and `REGION` is the ESSENCE Region,
a county name derived from a zip-code-to-county lookup table maintained by
ESSENCE (e.g., `"Jefferson"`). Some site names contain additional underscores
(e.g., multi-word state abbreviations used in certain NSSP configurations), so
the site prefix is defined as all characters before the **last** underscore in
the string.

Both geography functions use the same detection logic:
- A visit is classified as **in-state** when `Region` begins with
  `paste0(site, "_")` (e.g., `"KY_"` for `site = "KY"`) and is not the
  known-site unknown-residence placeholder below.
- A visit is classified as **out-of-state or unknown** when `Region` begins
  with a different prefix, is exactly `"OTHER_REGION"`, is exactly
  `paste0(site, "_UNKNOWN")` (e.g., `"KY_UNKNOWN"`), or is `NA`. The
  `"{site}_UNKNOWN"` form is checked explicitly rather than relying on the
  prefix mismatch, since it carries the site prefix and would otherwise be
  (incorrectly) classified as in-state.

```{r region-examples}
# Illustrate the detection logic on a few example values
region_examples <- c("KY_Jefferson", "TN_Davidson", "OH_Hamilton",
                     "OTHER_REGION", "KY_UNKNOWN", NA_character_)

data.frame(
  region   = region_examples,
  in_state = startsWith(
    tidyr::replace_na(region_examples, ""),
    "KY_"
  ) &
    !is.na(region_examples) &
    region_examples != "KY_UNKNOWN"
)
```

## Two Functions, Two Philosophies

`sysPrep` provides two geographic attribution functions that differ
fundamentally in **which rows are modified**:

| | `assign_treating_geography()` | `assign_facility_geography()` |
|---|---|---|
| **Default output column** | `region_hybrid` (new column; `Region` untouched) | `region_facility` (new column; `Region` untouched) |
| **Rows reassigned** | Out-of-state + `OTHER_REGION` + `NA` only | **All rows** |
| **In-state patient geography** | Reflected as-is in `region_hybrid` | Overwritten with facility geography |
| **Research question** | Where did this care take place? (for non-residents only) | Where did all care take place? |
| **Resulting column meaning** | Mixed: residential for in-state, treating for out-of-state | Treating facility county for everyone |

Both default to writing a new column rather than overwriting `Region` in
place; `Region` itself is never touched unless you explicitly opt into
that with `overwrite = TRUE` (covered at the end of this vignette). This
means calling both functions back to back gives you all three geographies
side by side with zero extra arguments:

```{r three-way-preview}
geo_all <- essence_raw |>
  dedupe(order_by = Arrived_Date_Time, keep = "last") |>
  filter_care_setting(
    fix_facility_type_vector = c("Hillside FSED", "Downtown Emergency Services")
  ) |>
  assign_treating_geography() |>
  assign_facility_geography()

geo_all |>
  dplyr::select(hospital_name, region, region_hybrid, region_facility) |>
  head(10)
```

`region` is the untouched original for every row; `region_hybrid` matches it
for in-state visits and substitutes treating geography for the rest;
`region_facility` substitutes treating geography for **every** row,
including the in-state majority. One pass over the data, three usable
geographies, no `preserve_original_geographies`, no manual copying.

This distinction matters for how the resulting data should be interpreted and
which analyses it supports.

## `assign_treating_geography()`: Selective Hybrid Reassignment

`assign_treating_geography()` is a **targeted intervention** on problem rows.
It identifies visits where the patient geography is unavailable or out-of-area
and writes the treating facility's county, for those visits only, to
`region_hybrid`. `region` is left exactly as received from ESSENCE for every
row, in-state and out-of-state alike.

`region_hybrid` is a **hybrid** column: it holds residential geography for
patients who provided an in-state address, and treating facility geography
for patients who could not or did not. This hybrid is appropriate for
area-based incidence analyses where you want to count as many visits as
possible within your surveillance area, and where the distinction between
resident and non-resident burden is not the primary analytical question.

```{r treating-geography}
# Start from deduplicated, filtered data
ed_clean <- essence_raw |>
  dedupe(order_by = Arrived_Date_Time, keep = "last") |>
  filter_care_setting(
    fix_facility_type_vector = c("Hillside FSED", "Downtown Emergency Services")
  )

# Selective reassignment: only out-of-state and OTHER_REGION rows get a
# reassigned value in region_hybrid; region itself is never touched
treating_geo <- ed_clean |>
  assign_treating_geography(site = "KY")
```

```{r treating-counts}
# How many visits were reassigned?
dplyr::count(treating_geo, .out_of_state)
```

```{r treating-comparison}
# Compare original region against the hybrid result for reassigned rows;
# no preserve_original_geographies needed, since region was never modified
treating_geo |>
  dplyr::filter(.out_of_state) |>
  dplyr::select(hospital_name, region, region_hybrid) |>
  head(10)
```

The `.out_of_state` column (logical) marks every row where `region_hybrid`
differs from `region`. Rows where `.out_of_state = FALSE` have
`region_hybrid == region`: the patient's residential county, unchanged.

If you'd rather overwrite `region`/`zip_code` in place instead of getting a
`region_hybrid`/`zip_code_hybrid` column (reproducing the only behavior
this function had before `new_region_col`/`new_zip_col` existed), pass
their own names explicitly and opt in with `overwrite = TRUE`:

```{r treating-overwrite}
# Opt-in in-place overwrite, with an audit trail of what was replaced
treating_overwrite <- ed_clean |>
  assign_treating_geography(
    site                           = "KY",
    new_region_col                 = "region",
    new_zip_col                    = "zip_code",
    overwrite                      = TRUE,
    preserve_original_geographies  = TRUE
  )

treating_overwrite |>
  dplyr::filter(.out_of_state) |>
  dplyr::select(hospital_name, original_region, region) |>
  head(10)
```

`preserve_original_geographies` only has an effect in this `overwrite = TRUE`
mode: when it applies, `original_region`/`original_zip_code` are added
before overwriting, retaining pre-reassignment values as an audit trail. For
in-state visits, `original_region` is `NA` because no change was made. In
the default (new-column) mode above, this parameter is unnecessary and has
no effect, since `region` was never modified in the first place.

**When to use `assign_treating_geography()`:**

- Incidence estimation where you want residential geography for most patients
  and facility geography only as a fallback for patients without valid
  residential addresses.
- Trend analyses where you need to retain border-facility visits without
  attributing all visits to facility location.
- Sensitivity analyses: compare results with and without `OTHER_REGION` visits
  using the `.out_of_state` flag to filter.

## `assign_facility_geography()`: Universal Full Reassignment

`assign_facility_geography()` takes a fundamentally different approach:
**every row** gets a facility-geography value in `region_facility`, regardless
of whether the patient was in-state, out-of-state, or unknown. `region`
itself is untouched, exactly as with `assign_treating_geography()`.
`region_facility` never reflects where patients live; it always reflects
where they received care.

```{r facility-geography}
# Universal reassignment: ALL rows get a facility geography value
facility_geo <- ed_clean |> assign_facility_geography()
```

```{r facility-comparison}
# region_facility vs. the untouched original, for every row
facility_geo |>
  dplyr::select(hospital_name, hospital_region, region, region_facility) |>
  head(10)
```

```{r facility-verify}
# Confirm: region_facility == hospital_region for every row
all(facility_geo$region_facility == facility_geo$hospital_region)
```

Unlike `assign_treating_geography()`, where only out-of-state/unknown rows
differ between `region` and `region_hybrid`, here **every** row differs
between `region` and `region_facility`; that's the "universal" part.

> **Is this only renaming `HospitalRegion` to `Region`?**
> Mechanically, close: `assign_facility_geography()` writes `HospitalRegion`
> and/or `HospitalZip` to `region_facility`/`zip_code_facility` for all rows,
> depending on which columns are present in the pull. Not all analysts
> include both; zip-based geography carries its own methodological
> considerations around spatial resolution and boundary alignment that
> may not be appropriate for every analysis. When only `HospitalRegion` is
> present, only region is reassigned. Some practitioners have applied this
> rename informally as an ad hoc pipeline step. `sysPrep` formalizes that
> practice: documents the intent, handles whichever geography types are
> available, and writes to a new column by default so the original values
> are always available for audit or comparison without any extra argument.
> `assign_treating_geography()`, the hybrid selective approach, has no
> simple column-rename equivalent, which is what motivated formalizing both
> methods in the same package.

Hospital-level spatial clustering (point-based) concentrates observations at
individual facility locations and can identify which specific facilities are
experiencing elevated burden, including spikes that would be diluted or
invisible at the regional level. However, hospital-level scan statistics
generally do not explicitly model residual spatial autocorrelation or
facility-specific reporting patterns as nuisance structure. In prospective
surveillance, scanning windows may therefore interact with hospital density,
catchment patterns, and data submission variability in ways that can produce
apparent clusters independent of true changes in underlying incidence. This
concern is more pronounced in urban areas, where hospitals are denser and
submission patterns may vary across nearby facilities. Reassigning all visits
to `HospitalRegion` before aggregation reduces facility-level noise by
aggregating across treating facilities within a region, allowing spacetime
permutation methods and other scan statistics to treat a rural county with one
hospital comparably to an urban county with many. Region-level binning does not
eliminate spatial autocorrelation concerns, but it reduces the influence of
facility-specific variability that is more common in high-density areas.

A second advantage is logistical: hospital-level spatial clustering requires
latitude/longitude coordinates from the NSSP Master Facility Table, an
additional lookup outside the standard ESSENCE data pull. Region-level
attribution achieves a close approximation of hospital-level cluster detection
using `HospitalRegion`, a field returned without any coordinate lookup,
**provided it's included in the pull**: `HospitalRegion` isn't guaranteed on
every ESSENCE pull by default, only when it's part of your site's default
returned columns or explicitly requested via the API's `&field=` parameter.
The same logistical advantage also supports a distinct, simpler use case:
total treated volume by urban vs. rural county, with no interest in which
specific hospital saw each visit at all, not an approximation of
hospital-level detection, just county-level counting using a field you
already requested rather than a coordinate lookup you'd have to add.

> **What does a county with no ESSENCE-participating hospital show?**
> `0`, always: not a small or uncertain number, a hard zero, regardless of
> how much true burden that county's residents actually generate. This is a
> direct consequence of ESSENCE being facility-based: `region_facility` can
> only ever reflect counties containing a facility that submits to ESSENCE.
> A county with real burden but no local ED will never contribute a count
> under this method, no matter how the underlying incidence changes. This
> is worth stating plainly rather than discovering it from a suspiciously
> flat map.

The two approaches are complementary: hospital-based clustering offers
facility-level resolution; region-based clustering offers geographic equity
across areas of differing hospital density. The choice depends on whether the
surveillance question is "which facility?" or "which area?".

**When to use `assign_facility_geography()`:**

- Analyses framed around treating facility location rather than patient
  residence: "How many cases were treated in Jefferson County?" rather than
  "How many residents of Jefferson County were treated?"
- Facility service area analyses where you want all visits attributed to the
  county where the facility is located.
- Spatial cluster detection scoped to treating location (e.g., identifying
  counties with high overdose treatment burden regardless of patient origin).
- Cross-state facility comparisons where residential geography is an
  inappropriate common denominator.
- Comparing total treated volume across urban and rural counties without
  differentiating which hospital saw each visit, useful when a
  facility-to-county lookup table isn't available but `HospitalRegion`/
  `HospitalZip` already are.

## The Core Distinction: What Do the Output Columns Mean?

The choice between the two functions is a choice about what geography your
analysis should be scoped to:

**`region_hybrid`** (from `assign_treating_geography()`):
residential geography for in-state patients + treating facility geography
for out-of-state and unknown-residence patients.

This is a **hybrid** with mixed semantics. It is most useful when you want to
maximize the number of visits attributed to your surveillance area while
preserving the residential geography of the majority.

**`region_facility`** (from `assign_facility_geography()`):
treating facility geography for all patients.

This is semantically **uniform**: `region_facility` means the same thing
for every row. It is most useful when the analytical unit is where care was
delivered, not where patients live.

`region` itself, meanwhile, is never touched by either function unless you
explicitly opt into `overwrite = TRUE`, so it's always available as the
untouched original, regardless of which (or both) of these you compute.
Neither function is universally correct. The choice should be documented in
your analysis methods and driven by the research question.

## Decision Guide

| Research question | Function | Rationale |
|-------------------|----------|-----------|
| County-level incidence rate (retain most OOS visits) | `assign_treating_geography()` | Preserves residential geography for in-state majority; uses facility geography as fallback |
| Which counties have highest treatment burden? | `assign_facility_geography()` | All visits attributed to treating county uniformly |
| Where do patients live who are treated at this facility? | Neither; use `Region` directly | Residential geography is the question; don't reassign |
| Spatial cluster detection by facility county | `assign_facility_geography()` | Uniform attribution aligns with facility-based cluster geometry |
| Sensitivity analysis: effect of OOS exclusion | `assign_treating_geography()` | Use `.out_of_state` flag to toggle OOS visits in/out |
| Trend analysis, mixed state border facility | `assign_treating_geography()` | Avoids flattening in-state residential variation |

## Output Columns: New by Default, Overwrite as an Opt-In

Every example above wrote to a new column (`region_hybrid`/`region_facility`)
and left `region`/`zip_code` alone. That's the default for a reason: it's
non-destructive, so calling either function is always safe to try, safe to
re-run, and safe to chain: there's no risk of losing the original values
just by calling the function.

`new_region_col`/`new_zip_col` control the target column and default to
`"region_hybrid"`/`"zip_code_hybrid"` (`assign_treating_geography()`) or
`"region_facility"`/`"zip_code_facility"` (`assign_facility_geography()`).
Pass `NULL` to either to skip that geography type entirely, e.g.
`new_zip_col = NULL` to reassign region only.

If the target you name already exists as a column, whether that's
`region_col`/`zip_col` itself, or any other existing column, the function
aborts rather than silently overwriting it, unless you set
`overwrite = TRUE`. This is what makes reproducing the original
in-place-overwrite behavior an explicit, visible choice rather than the
default:

```r
essence_clean |>
  assign_treating_geography(
    new_region_col = "region",
    new_zip_col    = "zip_code",
    overwrite      = TRUE
  )
```

`preserve_original_geographies = TRUE` only has an effect in this
`overwrite = TRUE` mode: it adds `original_region`/`original_zip_code`
columns holding the pre-overwrite values before they're replaced. In the
default new-column mode, it has no effect and is unnecessary, since `region`
was never modified in the first place; it already *is* the original.

For `assign_treating_geography()` in overwrite mode, `original_region`/
`original_zip_code` contain pre-reassignment values only for rows where
geography changed (`.out_of_state = TRUE`); for in-state visits they are
`NA`. For `assign_facility_geography()` in overwrite mode, they contain
pre-reassignment values for **every row**, because every row is reassigned.
