---
title: "Getting Started with sysPrep"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with sysPrep}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
library(sysPrep)
library(dplyr)
```

## The ESSENCE Preprocessing Problem

The National Syndromic Surveillance Program (NSSP) Electronic Surveillance
System for the Early Notification of Community-based Epidemics
([ESSENCE](https://www.cdc.gov/nssp/)) provides near-real-time access to
emergency department visit data from participating facilities across the United
States. Epidemiologists use ESSENCE to monitor disease trends, detect spatial
clusters, and evaluate the impact of public health interventions.

`sysPrep`'s preprocessing steps are **not required** to run these analyses --
many surveillance questions tolerate the raw data quality issues below without
materially changing conclusions. Applying them is most worthwhile for
**small-count, high-impact case definitions**, where minimizing
false-positive clusters/anomalies and preserving external validity matter
most (for example, before feeding visit-level data into SaTScan or a similar
spatial scan tool for cluster/anomaly detection). This is the context in
which these methods were developed: applied opioid overdose surveillance and
cluster detection using NSSP ESSENCE data.

Four categories of data quality problems are present in virtually every ESSENCE
pull, and can bias case counts or distort cluster/anomaly detection if left
unaddressed:

**Duplicate records.** A single clinical encounter may appear as multiple
rows in the same pull. Ordinary clinical updates (a lab result, a
discharge disposition) are correctly appended to the existing record with
no additional row. Duplication happens instead when a specific calculated
field disagrees between transmissions of what should be the same
encounter -- most often `C_Visit_Date_Time` (e.g., when a hospital updates
it as care progresses and an update crosses midnight) or
`C_Unique_Patient_ID` (e.g., when a temporary identifier is corrected to a
permanent one). Without deduplication, case counts are inflated and trends
are distorted.

**Non-emergency providers.** ESSENCE returns visits from any facility that
submitted a matching syndromic record, including primary care clinics and
medical specialty practices that are not equipped to treat emergency
presentations. Free-standing emergency departments (FSEDs), which are
functionally identical to hospital-based EDs, are sometimes onboarded to
ESSENCE with incorrect facility type codes. Both categories require explicit
handling before ED-based incidence can be estimated.

**Mis-submitted and invisible direct admissions.** A row showing
`HasBeenAdmitted = 1` and `HasBeenE = 0` can mean two different things, and
nothing in that row alone tells you which: a genuine direct admission with
no preceding ED visit, or an ED visit whose transition to inpatient care
wasn't recorded as one continuous encounter -- producing what looks, to
ESSENCE, like two unrelated visits for a single event. The former is
structurally invisible to standard `HasBeenE = 1` queries; the latter
double-counts a real encounter if the two records are combined without
linking them. Encounter linkage resolves both by combining ED and
inpatient pulls into unified care episodes.

**Out-of-state and unknown-residence visits.** ESSENCE records include all
patients treated at in-state facilities regardless of where they live.
Patients who cannot or do not provide an address are coded as `OTHER_REGION`.
Discarding these visits produces systematic underestimates of facility burden,
particularly at facilities near state borders or serving mobile populations.

`sysPrep` formalizes the preprocessing steps practitioners apply to address
each of these problems into a reproducible, documented R pipeline.

`sysPrep`'s functions have been validated against records pulled from the
NSSP ESSENCE `va_er` (Patient Location, Full Details) and `va_hosp`
(Facility Location, Full Details) data sources. Both sources return ED
visit and inpatient admission records, so all exported functions apply to
either. Column names and value formats referenced throughout this package
(e.g., `{SITE}_{REGION}` region strings, `HasBeenE`/`HasBeenAdmitted` flags)
reflect these two data sources.

## Installation

```{r install, eval = FALSE}
remotes::install_github("andrew-farrey/sysPrep")
```

## The Synthetic Dataset

`sysPrep` ships with `essence_raw`, a fully synthetic dataset that reproduces
all four categories of ESSENCE data quality problems.

```{r glimpse}
dplyr::glimpse(essence_raw)
```

The dataset contains approximately 193 rows and 18 columns. Key features
embedded in `essence_raw`:

- **13 duplicate rows** across four mechanisms: standard retransmissions,
  midnight-crossing date changes, patient identifier corrections, and one
  compound case combining two mechanisms simultaneously.
- **20 non-ED provider records** from a `"Primary Care"` and a
  `"Medical Specialty"` facility that should be excluded.
- **Two FSEDs** (`"Hillside FSED"` and `"Downtown Emergency Services"`)
  misclassified as `"Urgent Care"` -- they are valid ED providers but require
  facility type correction.
- **Out-of-state and `OTHER_REGION` records** in the `Region` column.

For full column documentation, see `?essence_raw`. `vignette("encounter-linkage")`
uses two smaller, dedicated datasets (`essence_ed_raw`/`essence_inp_raw`)
to demonstrate `link_encounters()`, since it needs two separately queried
ESSENCE pulls rather than the single pull `essence_raw` represents.

## The Recommended Pipeline

The preprocessing steps should be applied in the order shown below. Each
step depends on the output of the previous: deduplication first (to ensure
accurate row counts for downstream steps), care setting filtering second (to
remove non-ED providers before geographic attribution), and geography
assignment last (applied only to the valid ED population).

```{r pipeline}
clean <- essence_raw |>
  # Step 1: Remove duplicate records, retaining the most recently transmitted
  dedupe(order_by = Arrived_Date_Time, keep = "last") |>
  # Step 2: Filter to valid ED and inpatient providers; correct FSED types
  filter_care_setting(
    fix_facility_type_vector = c("Hillside FSED", "Downtown Emergency Services")
  ) |>
  # Step 3: Assign treating facility geography to out-of-state and OTHER_REGION visits
  assign_treating_geography()
```

**Step 1 -- Deduplication** (`dedupe()`): Groups records by facility × visit
identifier. Within each group, retains the single row with the latest
`Arrived_Date_Time` -- the most recently transmitted version of the record,
which is most likely to contain updated clinical information. The `order_by`
and `keep` arguments control this behavior and can be adjusted for different
surveillance contexts (see `?dedupe` and the deduplication vignette).

**Step 2 -- Care setting filtering** (`filter_care_setting()`): Retains only
facilities with `FacilityType` values consistent with emergency care. The
`fix_facility_type_vector` argument names facilities that should be treated
as emergency providers despite incorrect facility type codes -- in this case,
two FSEDs. Without this correction, those facilities' visits would be excluded
along with genuine non-ED providers.

**Step 3 -- Geographic attribution** (`assign_treating_geography()`): For
visits where `Region` is out-of-state or `OTHER_REGION`, writes the treating
facility's corresponding geographic fields to new `region_hybrid`/
`zip_code_hybrid` columns. `Region`/`ZipCode` themselves are never modified
-- for in-state visits, `region_hybrid`/`zip_code_hybrid` simply carry
through the original value unchanged.

Each of these functions reports what it did via informational console
messages -- useful during interactive exploration, but often unwanted when
the pipeline runs inside a rendered R Markdown report or other
non-interactive context. Pass `verbose = FALSE` to suppress these messages;
warnings and errors are always shown regardless, so real data quality issues
are never silently hidden:

```{r verbose-example, eval = FALSE}
clean_quiet <- essence_raw |>
  dedupe(order_by = Arrived_Date_Time, keep = "last") |>
  filter_care_setting(
    fix_facility_type_vector = c("Hillside FSED", "Downtown Emergency Services"),
    verbose = FALSE
  ) |>
  assign_treating_geography(verbose = FALSE)
```

## What Changed?

```{r compare}
cat("Raw rows:   ", nrow(essence_raw), "\n")
cat("Clean rows: ", nrow(clean), "\n")
cat("\nRows removed at each step:\n")
cat("  Duplicates removed:         ",
    nrow(essence_raw) - nrow(dedupe(essence_raw, order_by = Arrived_Date_Time)), "\n")

after_dedup <- essence_raw |>
  dedupe(order_by = Arrived_Date_Time, keep = "last")
after_filter <- after_dedup |>
  filter_care_setting(
    fix_facility_type_vector = c("Hillside FSED", "Downtown Emergency Services")
  )
cat("  Non-ED providers excluded:  ",
    nrow(after_dedup) - nrow(after_filter), "\n")
cat("\nNew columns added by assign_treating_geography():\n")
cat(" ", paste(setdiff(names(clean), names(essence_raw)), collapse = ", "), "\n")
```

```{r reassigned}
# How many visits had geography reassigned?
dplyr::count(clean, .out_of_state)
```

The `.out_of_state` column marks visits where `region_hybrid` differs from
`Region` -- i.e., where the treating facility's county was substituted for
an out-of-state or `OTHER_REGION` value. For in-state visits,
`region_hybrid` equals `Region` unchanged. Chaining
`assign_facility_geography()` onto the same pipeline adds a third column,
`region_facility`, with treating geography substituted for *every* visit --
useful for comparing residential, hybrid, and facility-based geography side
by side without touching `Region` at all. See
`vignette("geography-assignment")` for the full mechanics, including how to
overwrite `Region`/`ZipCode` in place instead if that's what your pipeline
needs.

## Function Reference

| Function | Purpose |
|----------|---------|
| `summarize_duplicates()` | Count duplicate records per facility and dataset-wide -- call before deduplication to understand data quality |
| `classify_duplicates()` | Classify the mechanism of each duplicate group (`visit_date_change`, `pid_change`, `patient_class_change`, or compound) |
| `dedupe()` | Remove duplicate records, retaining one row per facility × visit identifier |
| `filter_care_setting()` | Filter to valid ED/inpatient providers; correct misclassified FSEDs |
| `review_facility_ed_visits()` | Flag facilities with anomalous visit volumes using IQR or SD outlier detection |
| `link_encounters()` | Link ED and inpatient pulls into unified care episodes, correcting mis-submitted admissions and capturing genuine direct admissions |
| `assign_treating_geography()` | Reassign geography for out-of-state and `OTHER_REGION` visits only |
| `assign_facility_geography()` | Reassign geography for all visits to treating facility location |

## Next Steps

- `vignette("deduplication-and-classification")`: Detailed methodology for
  each duplication mechanism, how to interpret `summarize_duplicates()` and
  `classify_duplicates()` output, and a decision guide for choosing a
  deduplication strategy.

- `vignette("encounter-linkage")`: The care pathway artifact problem,
  `HasBeen_` column structure, linking a separately queried ED and
  admission pull, and burden estimation from multi-class episode data.

- `vignette("geography-assignment")`: When to use `assign_treating_geography()`
  versus `assign_facility_geography()`, the `Region` field format, and
  consequences of discarding out-of-state visits in facility-level analyses.
